Generating Random Logic Programs Using Constraint Programming

### 2020-01-25 paper/paper.tex
Freshly baked abstract

\begin{abstract} % TODO: rewrite
  We present a novel approach to generate random logic programs using constraint
  programming. The constraint programming approach to the problem has a major
  advantage in that one can easily add additional conditions for the generated
  programs. Some of these conditions can be expressed using constraints native to
  the constraint solver Choco \citep{choco}, while novel propagation and entailment
  checking algorithms have been developed for ensuring independence and/or
  conditional independence amongst predicates. A constraint programming approach
  also benefits from many years of research into efficient constraint propagation
  algorithms which makes it likely that our model can generate larger programs
  relatively quickly.
\end{abstract}

@@

% TODO: import related work from my proposal

Motivation:
\begin{itemize}
\item Empirical testing of inference algorithms.
\item Generating random programs that generate random data.
  %For example, enforcing that the decision is independent of gender.
\item Learning: how this can be used for (targeted) learning, when (atomic)
  probabilities can be assigned based on counting and we can have extra
  constraints. A more primitive angle: generate structures, learn weights.
\end{itemize}

%TODO: define all the relevant terminology from logic and constraint
% programming: predicates ,arities, (logic) variables, constants, clauses, atoms,
% heads, bodies, formulas.
% CP: decision variables, constraints. So the word 'variable' has two meanings.
% TODO: need to introduce the idea of predicate independence

We say that a constraint variable is \emph{(fully) determined} if its domain (at
the given moment in the execution) has exactly one value.

A \emph{(logic) program} is a multiset of clauses. Given a program
$\mathscr{P}$, a \emph{subprogram} $\mathscr{R}$ of $\mathscr{P}$ is a subset of
the clauses of $\mathscr{P}$ and is denoted by $\mathscr{R} \subseteq
\mathscr{P}$.

We will often use $\Box$ as a special domain value indicating a `disabled'
(i.e., fixed and ignored) part of the model. We write $\arrayd{a}{b}{c}$ to mean
that $\variable{a}$ is an array of variables of length $b$ such that each
element of $\variable{a}$ has domain $c$. Similarly, we write $\arrayt{a}{b}{c}$
to denote an array $\variable{a}$ of length $b$ such that each element of
$\variable{a}$ has type $\variable{c}$. All constraint variables in the model
are integer variables, but, e.g., if the integer $i$ refers to a logical
variable $X$, we will use $i$ and $X$ interchangeably. All indices start at
zero.

We also use Choco~4.10.2 \citep{choco}. This works with both Prolog
\citep{DBLP:books/daglib/0041598} and ProbLog \citep{DBLP:conf/ijcai/RaedtKT07}.
%Tested with SWI-Prolog \citep{DBLP:journals/tplp/WielemakerSTL12}.

@@

\section{HEADS OF CLAUSES}

@@

\section{COUNTING PROGRAMS}

@@

\begin{definition}
  Let $\mathsf{P}$ be a predicate in a program $\mathscr{P}$. The set of
  \emph{dependencies} of $\mathsf{P}$ is the smallest set $D_{\mathsf{P}}$ such
  that:
  \begin{itemize}
  \item $\mathsf{P} \in D_{\mathsf{P}}$,
  \item for every $\mathsf{Q} \in D_{\mathsf{P}}$, all direct predecessors of
    $\mathsf{Q}$ in $G_{\mathscr{P}}$ are in $D_{\mathsf{P}}$.
  \end{itemize}
\end{definition}

@@

\begin{algorithm}
  \LinesNumbered
  \SetKwFunction{getDependencies}{deps}
  \SetKwFunction{fail}{fail}
  \SetKwFunction{removeValue}{removeValue}
  \SetKwData{predicate}{predicate}
  \SetKwData{source}{source}
  \SetKwData{target}{target}
  \KwData{predicates $p_1$, $p_2$; adjacency matrix $\mathbf{A}$}
  \For{$(d_1, d_2) \in \getDependencies{$p_1$, 0} \times \getDependencies{$p_2$,
      0}$ such that $d_1.\predicate = d_2.\predicate$}{
    \lIf{$d_1$ {\bf is} $\Determined(\_)$ {\bf and} $d_2$ {\bf is}
      $\Determined(\_)$}{\fail{}} \label{line:fail}
    \If{$d_1$ {\bf is} $\Determined(\_)$ {\bf and} $d_2$ {\bf is}
      $\AlmostDetermined(\_, s, t)$ {\bf or} \hspace{50pt} $d_2$ {\bf is}
      $\Determined(\_)$ {\bf and} $d_1$ {\bf is} $\AlmostDetermined(\_, s,
      t)$ \label{line:propagation_condition}}{
      $\mathbf{A}[s][t]$.\removeValue{$1$}\; \label{line:propagation_result}
    }
  }
  \caption{Propagation for independence}
  \label{alg:independence_propagation}
\end{algorithm}

@@

\begin{example} \label{example:independence}
  Consider this partially determined (fragment of a) program:
  \begin{align*}
    \Box(X, Y) &\gets \mathsf{parent}(X, Z) \land \mathsf{parent}(Y, Z),\\
    \mathsf{father}(X, Y) &\gets \mathsf{parent}(X, Y) \land \neg\mathsf{mother}(X, Y)
  \end{align*}
  where $\Box$ indicates an unknown predicate with domain
  \[
    D = \{ \mathsf{father}, \mathsf{mother}, \mathsf{parent}, \mathsf{sibling}
    \}.
  \]
  The predicate dependency graph (without positivity/negativity)---as defined by
  \cref{def:adjacency_matrix}---is represented in
  \cref{fig:dependencies_matrix,fig:dependencies2}.

@@

\section{EMPIRICAL PERFORMANCE}

@@

In \cref{fig:impact}, we plot how the mean number of nodes in the binary search
tree grows as a function of each parameter (the plot for the median is very
similar). The growth of each curve suggest how well/poorly the model scales with
higher values of the parameter. From this plot, it is clear that
$\maxNumNodes{}$ is the limiting factor. This is because some tree structures
can be impossible to fill with predicates without creating either a negative
cycle or a forbidden dependence, and such trees become more common as the number
of nodes increases.

@@

% TODO: talk about randomness. The value heuristic is random because we want
% randomness. After finding one solution, one can continue the search for more,
% but they will be similar (acceptable if the search space is small and we want
% to find all solutions).
% On the other hand, with restarts, duplicates become possible.
% TODO: the number of predicates is also important for performance.

@@

% TODO: the positive side (what did we achieve?)

@@

>>>>

\begin{abstract}
  We present a novel approach to generating random logic programs and
  random probabilistic logic programs using constraint programming. The
  generated programs are useful in empirical testing of inference algorithms,
  random data generation, and program learning. This approach has a major
  advantage in that one can easily add additional conditions for the generated
  programs. As an example of this, we introduce a new constraint for predicate
  independence with efficient propagation and entailment algorithms, allowing
  one to generate programs that have a certain independence structure. In order
  to generate valid probabilistic logic programs, we also present a new
  constraint for negative cycle detection. Finally, we provide a combinatorial
  argument for correctness and describe how the parameters of the model affect
  the empirical difficulty of the program generation task.
\end{abstract}

@@

How confidently can we claim that an algorithm works well if it is only tested
on a couple types of problems? Perhaps the `inferior' algorithm is actually
better in some specific circumstances. Perhaps there are identifiable
types of inputs that make every algorithm default to exponential behaviour
that could be easily overcome with the right strategy. At present, most
inference algorithms for probabilistic logic programs are only evaluated on 1--4
different problems: sometimes just a single network
\citep{DBLP:journals/tplp/KimmigDRCR11,DBLP:journals/corr/abs-1112-3785,DBLP:conf/iclp/KimmigCRDR08},
sometimes two
\citep{DBLP:journals/corr/abs-1009-3798,DBLP:conf/ecai/BruynoogheMKGVJR10} or
four networks \citep{DBLP:conf/ijcai/VlasselaerBKMR15} coming from a range
of areas such as social networks and citation/genetic/biological data sets. In
order to better understand the strengths and weaknesses of these algorithms,
along with real data they should also be evaluated on a range of synthetic
problems that accurately represent the potential complexities that have to be
handled by a well-designed inference algorithm. Furthermore, random logic
program generators can be useful in combination with methods that generate
random data with probabilistic logic programs \citep{DBLP:conf/soict/Dries15}
and can be used as a component of learning.

Most current approaches to generating random logic programs are quite
restrictive, e.g., limited to clauses with only two literals
\citep{DBLP:conf/lpnmr/NamasivayamT09}, or to clauses of the form $a \gets \neg
b$ \citep{DBLP:journals/tocl/WenWSL16}, but others are more expressive, e.g.,
defining a program only by the (maximum) number of atoms in the body and the
total number of rules \citep{DBLP:conf/iclp/ZhaoL03,zhao2004answer}. We
introduce a way to generate random logic programs using a constraint solver such
as Choco \citep{choco}. The same model can generate both probabilistic programs
directly in the syntax of ProbLog \citep{DBLP:conf/ijcai/RaedtKT07} as well as
non-probabilistic Prolog programs. For generated probabilistic programs to be
valid, we use a custom constraint to detect negative cycles (as described in
\cref{sec:cycles}). A major advantage of our constraint-based approach is that
one can easily add additional constraints to the model. To demonstrate that, in
\cref{sec:independence} we present a custom constraint with propagation and
entailment algorithms that can ensure predicate independence.
\cref{sec:counting} also presents a combinatorial argument for correctness,
where we produce combinatorial expressions that count the number of programs
that the model should produce for various parameter values. Finally,
\cref{sec:experiments} shows how the model scales when tasked with producing
more complicated programs and identifies the relationships between parameter
values and the empirical hardness of the program generation task.

\subsection{PRELIMINARIES}

The basic primitives of logic programs are \emph{constants}, \emph{(logic)
  variables}, and \emph{predicates}. Each predicate has an \emph{arity} that
defines the number of terms that it can applied to. A \emph{term} is either a
variable or a constant, and an \emph{atom} is a predicate of arity $n$ applied
to $n$ terms. A \emph{formula} is a grammatically-valid expression that connects
atoms using conjunction ($\land$), disjunction ($\lor$), and negation ($\neg$).
A \emph{clause} is a pair of a \emph{head} (which is an atom) and a \emph{body}
(which is a formula). A \emph{(logic) program} is a multiset of clauses. Given a
program $\mathscr{P}$, a \emph{subprogram} $\mathscr{R}$ of $\mathscr{P}$ is a
subset of the clauses of $\mathscr{P}$ and is denoted by $\mathscr{R} \subseteq
\mathscr{P}$.

In the world of constraint satisfaction, we also have \emph{(constraint)
  variables}, each with its own \emph{domain}, whose values are restricted using
\emph{constraints}. All constraint variables in the model are integer or set
variables, however, if an integer refers to a logical construct
(e.g., a logical variable or a constant), we will make no distinction between
the two and often use names of logical constructs to refer to the underlying
integers. We say that a constraint variable is \emph{(fully) determined} if its
domain (at the given moment in the execution) has exactly one value. We will
often use $\Box$ as a special domain value to indicate a `disabled' (i.e., fixed
and ignored) part of the model. We write $\arrayd{a}{b}{c}$ to mean that
$\variable{a}$ is an array of variables of length $b$ such that each element of
$\variable{a}$ has domain $c$. Similarly, we write $\arrayt{a}{b}{c}$ to denote
an array $\variable{a}$ of length $b$ such that each element of $\variable{a}$
has type $\variable{c}$. Finally, we assume that all arrays start with index
zero.

@@

\section{HEADS OF CLAUSES} \label{sec:heads}

@@

\section{COUNTING PROGRAMS} \label{sec:counting}

@@

\begin{definition}
  Let $\mathsf{P}$ be a predicate in a program $\mathscr{P}$. The set of
  \emph{dependencies} of $\mathsf{P}$ is the smallest set $D_{\mathsf{P}}$ such
  that $\mathsf{P} \in D_{\mathsf{P}}$, and, for every $\mathsf{Q} \in
  D_{\mathsf{P}}$, all direct predecessors of $\mathsf{Q}$ in $G_{\mathscr{P}}$
  are in $D_{\mathsf{P}}$.
\end{definition}

@@

\begin{algorithm}[h]
  \LinesNumbered
  \SetKwFunction{getDependencies}{deps}
  \SetKwFunction{fail}{fail}
  \SetKwFunction{removeValue}{removeValue}
  \SetKwData{predicate}{predicate}
  \SetKwData{source}{source}
  \SetKwData{target}{target}
  \KwData{predicates $p_1$, $p_2$; adjacency matrix $\mathbf{A}$}
  \For{$(d_1, d_2) \in \getDependencies{$p_1$, 0} \times \getDependencies{$p_2$,
      0}$ such that $d_1.\predicate = d_2.\predicate$}{
    \lIf{$d_1$ {\bf is} $\Determined(\_)$ {\bf and} $d_2$ {\bf is}
      $\Determined(\_)$}{\fail{}} \label{line:fail}
    \If{$d_1$ {\bf is} $\Determined(\_)$ {\bf and} $d_2$ {\bf is}
      $\AlmostDetermined(\_, s, t)$ {\bf or} \hspace{50pt} $d_2$ {\bf is}
      $\Determined(\_)$ {\bf and} $d_1$ {\bf is} $\AlmostDetermined(\_, s,
      t)$ \label{line:propagation_condition}}{
      $\mathbf{A}[s][t]$.\removeValue{$1$}\; \label{line:propagation_result}
    }
  }
  \caption{Propagation for independence}
  \label{alg:independence_propagation}
\end{algorithm}

@@

\begin{example} \label{example:independence}
  Consider this partially determined (fragment of a) program:
  \begin{align*}
    \Box(X, Y) &\gets \mathsf{parent}(X, Z) \land \mathsf{parent}(Y, Z),\\
    \mathsf{father}(X, Y) &\gets \mathsf{parent}(X, Y) \land \neg\mathsf{mother}(X, Y)
  \end{align*}
  where $\Box$ indicates an unknown predicate with domain
  \[
    D = \{ \mathsf{father}, \mathsf{mother}, \mathsf{parent}, \mathsf{sibling}
    \}.
  \]
  The predicate dependency graph without positivity/negativity (as defined in
  \cref{def:adjacency_matrix}) is represented in
  \cref{fig:dependencies_matrix,fig:dependencies2}.

@@

\section{EMPIRICAL PERFORMANCE} \label{sec:experiments}

@@

In \cref{fig:impact}, we plot how the mean number of nodes in the binary search
tree grows as a function of each parameter (the plot for the median is very
similar). The growth of each curve suggest how well/poorly the model scales with
higher values of the parameter. From this plot, it is clear that
$\maxNumNodes{}$ is the limiting factor. This is because some tree structures
can be impossible to fill with predicates without creating either a negative
cycle or a forbidden dependence, and such trees become more common as the number
of nodes increases. Likewise, a higher number of predicates complicates the
situation as well.

@@

% TODO: talk about randomness. The value heuristic is random because we want
% randomness. After finding one solution, one can continue the search for more,
% but they will be similar (acceptable if the search space is small and we want
% to find all solutions).
% On the other hand, with restarts, duplicates become possible.

@@

% TODO: the positive side (what did we achieve?)
%We were able to design an efficient model

@@

\newpage
